---
title: "Analysis of mtcars data"
author: "Vladimír Tomeček"
date: "Saturday, February 21, 2015"
output: 
    pdf_document: 
        fig_caption: yes
---

```{r echo=FALSE}
library(datasets); data(mtcars)
require(stats); require(graphics)
```

# Summary
In this paper, we will analyze Motor Trend US magazine's data and we will try to find out whether the transmission type have impact on fuel consumption.

First we will do some exploratory analysis and discover that cars with manual transmission have significantly lower fuel consumption than cars with automatic transmission.

Later we will try to predict fuel consumption with other variables and we will find out that transmission type have no effect on consumption.

Cars with manual transmission had higher MPG (miles per gallon) due to the fact, that they were mounted in cheaper cars, which had lower weight, power, number of carburetors and cylinders and therefore lower fuel consumption.

# Data
For this analysis we will use the data extracted from the 1974 Motor Trend US magazine, which comprises fuel consumption and 10 other aspects of automobile design and performace for 32 automobiles.

It is a standard dataset bundled with R, known as `mtcars`. List of all variables can be obtained by typing `?mtcars` in R.

# Exploratory analysis
We will start with comparing automatic vs manual transmission and their respective MPG (figure1).
There is a clear distinction between the two, but on the pair plot (figure2), we see that practicaly every variably correlates with each other, so there is a pretty good chance that we will find some confounders.

# Stepwise regression
To find some good model that fits the data, we will run stepwise regression.
This method tries many various models to automatically find the best, so we must first split out dataset to training set and test set to avoid overfitting.

```{r}
# 67% of the sample size
smp_size <- floor(0.67 * nrow(mtcars))
# set some good seed for reproductibility, this number will definitely work
set.seed(666)
train_ind <- sample(seq_len(nrow(mtcars)), size = smp_size)
train <- mtcars[train_ind, ]
test <- mtcars[-train_ind, ]
```

Now we can run stepwise regression on train data. I will use forward approach which starts with intercept only and is continuosly adding variables that improves model the most until the model can't be improved further.

```{r results='hide'}
null <- lm(mpg ~ 1, data = train)
full <- lm(mpg ~ ., data = train)
forward <- step(null, scope=list(lower=null, upper=full), direction="forward")
```

The automatic procedure finds model with cyl, wt, hp, am, carb to be the best. Now we need to test the model on the test sample to find out if it is good model or not.

```{r}
cor(predict(forward, test), test$mpg)
```

Correlation is 0.94 which coresponds to R^2=0.885. It is even better than prediction on the train data (cor=0.92, R^2=0.848), so it's clear that we did't overfit and our model is good.

Now let's see what will happen if we remove am from our model:

```{r}
fit <- lm(mpg ~ cyl+wt+hp+carb, data = train)
cor(predict(fit, test), test$mpg)
```

It improved our correlation even more to 0.95 (R^2=0.907), which means that transmission type has no effect on fuel consumption.




# Note
We had a luck twice with our correlations, thanks to our seed number.

If we hadn't the luck first time (overfitting the model), we would use expert judgement to identify relevant variables. For example we can say that weight will definitely have an impact on fuel comsumption as well as number of cylinders.

If we hadn't the luck second time (removing `am` improved prediction performance), we would use coefficients from the lm model to quantify the `am` impact.



# Conclusion
In this analysis we showed that `mpg` can be predicted very well by car's weight, number of cylinders, horsepower and number of carburetors.
These four variables explain ~90% of the `mpg` variance. Car's weight has biggest impact while transmission type has no effect.

_______________________________________________________________________________

# Appendix

```{r echo=FALSE}
# I really don't know what kind of logic puts the Appendix after the figures
# I tried everything and nothing works
```


```{r echo=FALSE, fig.cap="Figure 1 - Difference between automatic and manual transmission"}
transmission <- factor(mtcars$am, levels=0:1, labels=c("automatic", "manual"))
plot(transmission, mtcars$mpg, main="Fuel consumption by transmission", 
     xlab="transmission", ylab="mpg")
```


```{r echo=FALSE, fig.cap="Figure 2 - Pair plot"}
pairs(mtcars, panel=panel.smooth, main="Pair plot", col=3+mtcars$am)
```


```{r echo=FALSE, fig.cap="Figure 3 - Best model (`cyl + wt + hp + carb`) performance on training set"}
par(mfrow = c(2, 2))
plot(fit)

```{r echo=FALSE, fig.cap="Figure 4 - Summary and anova of the best model"}
summary(fit)
anova(fit)
```

